A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases

### 2021-08-26 submission/manuscript.rmd
add responses to reviewer 1
more response work on reviewers 1 and 2

\newpage
\newpage
# Abstract {-}
\singlespacing 
One of the aims of the Observational Health Data Sciences and Informatics (OHDSI)
initiative is population-level treatment effect estimation in large
observational databases. Since treatment effects are well-known to vary across
groups of patients with different baseline risk, we aimed to extend the OHDSI
methods library with a framework for risk-based assessment of treatment effect
heterogeneity. The proposed framework consists of five steps: 1) definition of
the problem, i.e. the population, the treatment, the comparator and the
outcome(s) of interest; 2) identification of relevant databases; 3) development
of a prediction model for the outcome(s) of interest; 4) estimation of
propensity scores within strata of predicted risk and estimation of relative and
absolute treatment effect within strata of predicted risk; 5) evaluation and
presentation of results. We demonstrate our framework by evaluating
heterogeneity of the effect of angiotensin-converting enzyme (ACE) inhibitors
versus beta blockers on a set of 9 outcomes of interest across three
observational databases. With increasing risk of acute myocardial infarction we
observed increasing absolute benefits, i.e. from -0.03% to 0.54% in the lowest
to highest risk groups. Cough-related absolute harms decreased from 4.1%
to 2.6%. The proposed framework may be useful for the evaluation of heterogeneity
of treatment effect on observational data that are mapped to the OMOP Common
Data Model. The proof of concept study demonstrates its feasibility in large
observational data. Further insights may arise by application to safety and
effectiveness questions across the global data network.

@@

The proposed framework defines 5 distinct steps that enable a standardized
approach for risk-based assessment of treatment effect heterogeneity for
databases mapped to the OMOP-CDM. These are: 1) general definition of the
research aim; 2) identification of the database within which the analyses will
be performed; 3) a prediction step where internal or external prediction models
are used to assign patient-level risk predictions; 4) an estimation step where
absolute and relative treatment effects are estimated within risk strata; 5)
presentation and evaluation of the results. An overview of the procedure
can be seen in Figure \@ref(fig:figure1).

@@

The typical research aim is: "to compare the effect of treatment $T$ to a
comparator treatment $C$ in patients with disease $D$ with respect to outcomes
$O_1,\dots,O_n$". At least three cohorts are defined:

@@

Note "cohort" refers to a set of persons who satisfy one or more criteria for a
duration of time. The term can be used interchangeably with the term phenotype.

@@

The aim of this step is the inclusion of databases that represent the patient
population of interest. The inclusion of multiple databases potentially
increases the generalizability of results. Furthermore, the cohorts should
preferably have adequate sample size to ensure precise effect estimation, even
within smaller risk strata.

@@

We adopt the standardized framework for the generation of patient-level
prediction models using observational data that ensures adherence to existing
guidelines [@Collins2015; @Moons2015]. This prediction framework requires the
definition of two essential cohorts: a target cohort and an outcome cohort.

To generate the target cohort we pool the already defined treatment cohort and
comparator cohort. To avoid deferentially fitting the prediction model to
patients across treatment arms, thus introducing spurious interactions with
treatment [@Burke2014; @vanKlaveren2019], we develop the patient-level
prediction model in the propensity score-matched (1:1) subset of the population.

@@

***More specifically, we first estimated propensity scores using LASSO
logistic regression and a large set of baseline covariates including
demographics, drug exposures, diagnoses, measurements and medical devices.***

@@

Externally derived and well transportable prediction models are preferred for
analyzing treatment effect heterogeneity [@Kent2010]. In the absence of such
prediction models, simulations of RCTs have shown that internal models can be
used to provide unbiased estimates of treatment effect across the spectrum of
baseline risk [@Burke2014]. However, in observational databases treatment arms
may significantly differ in sample size. Because the prediction model will
possibly better fit to the larger treatment arm, spurious treatment-covariate
interactions may be introduced in the prediction model, leading to sub-optimal
risk stratification. As a remedy, we first match the patients in the treatment
and the comparator cohorts on the basis of propensity scores. Additionally, we
propose to assess model performance in the separate treatment arms to evaluate
its aptness for risk stratification.

>>>>

\newpage
\newpage
# Abstract {-}
\singlespacing 
One of the aims of the Observational Health Data Sciences and Informatics
(OHDSI) initiative is population-level treatment effect estimation in large
observational databases. Since treatment effects are well-known to vary across
groups of patients with different baseline risk, we aimed to extend the OHDSI
library of open-source tools with a framework for risk-based assessment of
treatment effect heterogeneity. The proposed framework consists of five
steps: 1) definition of the problem, i.e. the population, the treatment, the
comparator and the outcome(s) of interest; 2) identification of relevant
databases; 3) development of a prediction model for the outcome(s) of
interest; 4) estimation of propensity scores within strata of predicted risk and
estimation of relative and absolute treatment effect within strata of predicted
risk; 5) evaluation and presentation of results. We demonstrate our framework by
evaluating heterogeneity of the effect of angiotensin-converting enzyme (ACE)
inhibitors versus beta blockers on a set of 9 outcomes of interest across three
observational databases. With increasing risk of acute myocardial infarction we
observed increasing absolute benefits, i.e. from -0.03% to 0.54% in the lowest
to highest risk groups. Cough-related absolute harms decreased from 4.1%
to 2.6%. The proposed framework may be useful for the evaluation of
heterogeneity of treatment effect on observational data that are mapped to the
OMOP Common Data Model. The proof of concept study demonstrates its feasibility
in large observational data. Further insights may arise by application to safety
and effectiveness questions across the global data network.

@@

The proposed framework defines 5 distinct steps that enable a standardized
approach for risk-based assessment of treatment effect heterogeneity for
databases mapped to the OMOP-CDM. These are: 1) general definition of the
research aim; 2) identification of the database within which the analyses will
be performed; 3) a prediction step where internal or external prediction models
are used to assign patient-level risk predictions; 4) an estimation step where
absolute and relative treatment effects are estimated within risk strata; 5)
presentation and evaluation of the results. We developed an open-source
R-package for the implementation of the proposed framework and made it publicly
available. The source code can be found at
<https://github.com/OHDSI/RiskStratifiedEstimation>. An overview of the
procedure can be seen in Figure \@ref(fig:figure1).

@@

The typical research aim is: "to compare the effect of treatment $T$ to a
comparator treatment $C$ in patients with disease $D$ with respect to outcomes
$O_1,\dots,O_n$".

***Cohort definitions are crucial for this step of the framework. We define a
cohort as the set of patients who satisfy one or more inclusion criteria for a
duration of time. A cohort within the OHDSI setting is more than a set of
specific clinical codes, providing a definition of a logic for how to use that
code set. All cohort definitions consist of: an entry event, i.e. the time a
patient enters a cohort; a set of inclusion criteria applied to the initial
event cohort to further restrict the set of people, resulting in the
construction of the construction of the qualifying cohort; cohort exit criteria
that terminate the patient's presence in the cohort. Cohort definitions are
transportable, meaning they can be implemented in any database, provided that it
is mapped to the OMOP-CDM.***

@@

@@

The aim of this step is the inclusion of databases that represent the patient
population of interest. It is required that the databases are mapped to the
OMOP-CDM. The inclusion of multiple databases potentially increases the
generalizability of results. Furthermore, the cohorts should preferably have
adequate sample size to ensure precise effect estimation, even within smaller
risk strata.

@@

We adopt the standardized framework for the generation of patient-level
prediction models using observational data that ensures adherence to existing
guidelines [@Collins2015; @Moons2015]. This prediction framework requires the
definition of two essential cohorts: a target cohort, i.e. a set of patients
that satisfy one or more inclusion criteria for a duration of time, and an
outcome cohort.

To generate the target cohort we pool the already defined treatment cohort and
comparator cohort. Further restrictions can be applied on the target cohort to
construct the final population on which the prediction model will be developed
(e.g. exclude patients with a prior outcome in their history, before being
included in the target cohort). To avoid deferentially fitting the prediction
model to patients across treatment arms, thus introducing spurious interactions
with treatment [@Burke2014; @vanKlaveren2019], we develop the patient-level
prediction model in the propensity score-matched (1:1) subset of the population.

@@

***More specifically, we first estimate propensity scores using LASSO
logistic regression and a large set of baseline covariates including
demographics, drug exposures, diagnoses, measurements and medical devices. We
match patients 1-1 using a caliper, i.e. the maximum distance that is acceptable
for any match. The default value we use is 0.2 on the standardized logit scale
for the propensity scores.***

@@

***Ideally, externally derived and adequately validated prediction model would
be preferred for analyzing treatment effect heterogeneity [@Kent2010]. In the
absence of such prediction models an internally-developed risk prediction model
can be considered. Earlier simulations of RCT studies have shown that internal
models developed on the combined treatment and control arms blinded to treatment
gave relatively unbiased estimates of treatment effect across the spectrum of
risk [@Burke2014].***

Externally derived and well transportable prediction models are preferred for
analyzing treatment effect heterogeneity [@Kent2010]. In the absence of such
prediction models, simulations of RCTs have shown that internal models can be
used to provide unbiased estimates of treatment effect across the spectrum of
baseline risk [@Burke2014].

However, in observational databases treatment arms
may significantly differ in sample size. Because the prediction model will
possibly better fit to the larger treatment arm, spurious treatment-covariate
interactions may be introduced in the prediction model, leading to sub-optimal
risk stratification. As a remedy, we first match the patients in the treatment
and the comparator cohorts on the basis of propensity scores. Additionally, we
propose to assess model performance in the separate treatment arms to evaluate
its aptness for risk stratification.
